Turbo fast no compromise gravitational wave parameter estimation

### 2023-01-12 src/tex/ms.tex
update citation

% Problem with just HMC or MALA
While gradient-based samplers have been shown to be superior in terms of
performance when compared to other gradient-free algorithms in many practical
examples , there are still a number of classes of problems
gradient-based samplers do not solve well. \footnote{To be specific, we are
referring to sampling algorithms that use first-order derivatives here. Sampling
algorithms that use the information of higher order derivatives such as manifold-MALA
and Riemannian-HMC \cite{RMHMC} can in principle decorrelate local correlation in the target
distribution, however, they often have instability issues when they are used on
real-life applications, so they are not used often in practice.} For example,
target distribution that exhibits local correlation is hard to deal with,
since by construction first order gradient-based algorithms can only handle
global correlation structure. Another example is multi-modality. If there are
multiple modes in the target distribution, an independent chain will likely 
be trapped in one mode and take an extremely long time to transverse between the
modes. This means the relative weights between modes will much longer to sample
compared to the shape of each mode.

@@

% Crux of normalizing flow
Normalizing flow is a neural-network based technique that aims at learning a
mapping between a simple distribution, such as a Gaussian distribution, to a
complex distribution, often given in the form of data samples from a target
distribution. Once the network is trained, one can evaluate the probability
density of the complex distribution and samples from it very efficiently.
The core of normalizing flow is given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|,
\end{align}
where $p_x(X)$ is the complex target distribution $p_z(Z)$ is the simple latent
distribution and $f$ is an invertible parameterized transform that
connect the two distributions. For a detailed discussion of the algorithm, we
refer the readers to \kw{cite}.

% The flowMC algorithm
As mentioned in the previous subsection, one of the main issue gradient-based
sampler is it does not explore locally correlated features and multi-modality well. This
is exactly where the normalizing flow model can help. Once multiple independent
chains have generated samples, we can combine the samples and feed them to the
normalizing flow, which learns the global landscape of the target
distribution. Note that since the independent chains are generated with the same
number of samples, the normalization between posterior density represented by
different chains is in general different from the true target distribution. To
adjust for this inaccuracy, we use the normalizing flow model as a
\textit{proposal} distribution. Given a trained normalizing flow model, we can
generate the proposed jump in the target space by sampling from the latent
distribution $z \sim p_z(Z)$, then push it through the learned map given by the
normalizing flow model $x=f(z)$. the acceptance criteria is given by
\begin{align}
    \alpha(\mathbf{x},\mathbf{x_0}) = \min \left[ 1, \frac{\hat{\rho}(\mathbf{x_0})\rho_*(\mathbf{x})}{\hat{\rho}(\mathbf{x})\rho_*(\mathbf{x_0})}\right],
\end{align}
where $\hat{\rho}$ is the probability density estimated by the normalizing flow
model, $\rho_*$ is the probability density evaluated using the target function,
and $x_0$ is the current position. We can see the flow distribution is the
target distribution when the accepting probability is 1. When the normalizing
flow model has not converged to the target distribution, only a portion of the
proposed jumps will be accepted. This means an MCMC process using the
normalizing flow model as the proposal distribution can adjust the normalization
across different regions of the target parameter space by rejecting jumps into
less likely regions. The training and sampling are then repeated until certain
criteria are met. Note that every time we retrain the network, we are breaking
the Markov properties since we are changing the proposal distribution. To
produce final samples that can be used to estimate target quantities, one has to
freeze the normalizing flow model and not retrain during the final sampling
phase in order to satisfy the detailed balance condition. To summarize the
sampling process, the pseudocode is given by

>>>>

% Problem with just HMC or MALA
While gradient-based samplers have been shown to be superior in terms of
performance when compared to other gradient-free algorithms in many practical
examples, there are still a number of classes of problems gradient-based
samplers do not solve well. \footnote{To be specific, we are referring to
sampling algorithms that use first-order derivatives here. Sampling algorithms
that use the information of higher order derivatives such as manifold-MALA and
Riemannian-HMC \cite{RMHMC} can in principle decorrelate local correlation in
the target distribution, however, they often have instability issues when they
are used on real-life applications, so they are not used often in practice.} For
example, target distribution that exhibits local correlation is hard to deal
with, since by construction first order gradient-based algorithms can only
handle global correlation structure \cite{2017arXiv170102434B}. Another example
is multi-modality. If there are multiple modes in the target distribution, an
independent chain will likely be trapped in one mode and take an extremely long
time to transverse between the modes \cite{2018arXiv180803230M}. This means the
relative weights between modes will much longer to sample compared to the shape
of each mode.

@@

% Crux of normalizing flow
Normalizing flow is a neural-network based technique that aims at learning a
mapping between a simple distribution, such as a Gaussian distribution, to a
complex distribution, often given in the form of data samples from a target
distribution \cite{2019arXiv190809257K, 2019arXiv191202762P}. Once the network is trained, one can evaluate the probability
density of the complex distribution and samples from it very efficiently.
The core equation of normalizing flow is given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|,
\end{align}
where $p_x(X)$ is the complex target distribution $p_z(Z)$ is the simple latent
distribution and $f$ is an invertible parameterized transform that
connect the two distributions. For a detailed discussion of the algorithm, we
refer the readers to \kw{cite}.

% The flowMC algorithm
As mentioned in the previous subsection, one of the main issue gradient-based
sampler is it does not explore locally correlated features and multi-modality well. This
is exactly where the normalizing flow model can help. Once multiple independent
chains have generated samples, we can combine the samples and feed them to the
normalizing flow, which learns the global landscape of the target
distribution. Note that since the independent chains are generated with the same
number of samples, the normalization between posterior density represented by
different chains is in general different from the true target distribution. To
adjust for this inaccuracy, we use the normalizing flow model as a
\textit{proposal} distribution. Given a trained normalizing flow model, we can
generate the proposed jump in the target space by sampling from the latent
distribution $z \sim p_z(Z)$, then push it through the learned map given by the
normalizing flow model $x=f(z)$. The acceptance criterion is given by
\begin{align}
    \alpha(\mathbf{x},\mathbf{x_0}) = \min \left[ 1, \frac{\hat{\rho}(\mathbf{x_0})\rho_*(\mathbf{x})}{\hat{\rho}(\mathbf{x})\rho_*(\mathbf{x_0})}\right],
\end{align}
where $\hat{\rho}$ is the probability density estimated by the normalizing flow
model, $\rho_*$ is the probability density evaluated using the target function,
and $x_0$ is the current position. We can see the flow distribution is the
target distribution when the accepting probability is 1. When the normalizing
flow model has not converged to the target distribution, only a portion of the
proposed jumps will be accepted. This means an MCMC process using the
normalizing flow model as the proposal distribution can adjust the normalization
across different regions of the target parameter space by rejecting jumps into
less likely regions. The training and sampling are then repeated until certain
criteria are met. Note that every time we retrain the network, we are breaking
the Markov properties since we are changing the proposal distribution. To
produce final samples that can be used to estimate target quantities, one has to
freeze the normalizing flow model and not retrain during the final sampling
phase in order to satisfy the detailed balance condition. We use the package
\texttt{flowMC} \cite{2022arXiv221106397W}. The pseudocode of the algorithm is
given in Algorithm \ref{alg:cap}.
